Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-05-05 paper.qmd
Updated Figure 7

mse_two_curvy_b <- ggplot(error_two_curvy_umap, 
                     aes(x = a1, 
                         y = MSE)) +
  geom_vline(xintercept = 0.09,
           color = "#d95f02", linetype=2, 
           linewidth=1) +
  geom_vline(xintercept = 0.06,
             color = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype=2,
             color = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) + 
  geom_point(size = 1) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_umap$a1, 2)))[c(2, 5, 8, 10, 12, 14, 15, 16, 17)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = "MSE") +
  ggtitle("(a)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))

prop_dens_a1 <- ggplot(error_two_curvy_umap,
                     aes(x = a1,
                         y = prop_comp)) +
  geom_vline(xintercept = 0.09,
           color = "#d95f02", linetype=2, 
           linewidth=1) +
  geom_vline(xintercept = 0.06,
             color = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype=2,
             color = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_umap$a1, 2)))[c(2, 5, 8, 10, 12, 14, 15, 16, 17)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = paste("Relative proportion density")) +
  ggtitle("(d)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 15),
axis.text = element_text(size = 12))

@@

a1_m_two_curvy <- ggplot(error_two_curvy_umap,
                     aes(x = a1,
                         y = prop_bins)) +
  geom_vline(xintercept = 0.09,
           color = "#d95f02", linetype=2, 
           linewidth=1) +
  geom_vline(xintercept = 0.06,
             color = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype=2,
             color = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_umap$a1, 2)))[c(2, 5, 8, 10, 12, 14, 15, 16, 17)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = expression(paste("proportion of non-empty bins ", bgroup("(", frac(m, b), ")")))) +
  ggtitle("(b)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 15),
axis.text = element_text(size = 12))
```

@@

```{r}
#| echo: false
#| fig-cap: "Various plots to help assess best hexagon bin parameters (a, b, d) and thresholds to remove low-density bins (c). Both (a) and (c) show MSE, against binwidth ($a_1$) and threshold. A good benchmark value for these parameters is when the MSE drops and then flattens out. Three binwidth choices were made: $0.03$, $0.06$, and $0.09$ to investigate. As the binwidth increases, the proportion of non-empty bins also increases (b). The relative proportion density decreases and levels off (d). Binwidth $0.06$ is chosen as the initial best binwidth for further analysis. There is no need to remove the low-density hexagons because as shown in (b), there is no considerable drop in MSE."
#| label: fig-param-two-curvy
#| out-width: 80%
#| fig-width: 12
#| fig-height: 10
#| fig-pos: H

@@

## hexagon binning to have regular hexagons
hb_obj_tsne_two_curvy2 <- hex_binning(
  data = tsne_two_curvy_scaled, 
  bin1 = 8, 
  r2 = r2_tsne)

a1_2_tsne <- calc_bins_y(
  bin1 = 8, 
  r2 = r2_tsne
)$a1

>>>>

mse_two_curvy_b <- ggplot(error_two_curvy_umap, 
                     aes(x = a1, 
                         y = MSE)) +
  geom_vline(xintercept = 0.09,
           color = "#d95f02", linetype=2, 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             color = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype=2,
             color = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) + 
  geom_point(size = 1) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_umap$a1, 2)))[c(2, 4, 8, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = "MSE") +
  ggtitle("(a)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))

prop_dens_a1 <- ggplot(error_two_curvy_umap,
                     aes(x = a1,
                         y = prop_comp)) +
  geom_vline(xintercept = 0.09,
           color = "#d95f02", linetype=2, 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             color = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype=2,
             color = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_umap$a1, 2)))[c(2, 4, 8, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = paste("Relative proportion density")) +
  ggtitle("(d)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 15),
axis.text = element_text(size = 12))

@@

a1_m_two_curvy <- ggplot(error_two_curvy_umap,
                     aes(x = a1,
                         y = prop_bins)) +
  geom_vline(xintercept = 0.09,
           color = "#d95f02", linetype=2, 
           linewidth=1) +
  geom_vline(xintercept = 0.05,
             color = "#1f78b4", 
             linetype="solid",
             linewidth=1) +
  geom_vline(xintercept = 0.03, linetype=2,
             color = "#1b9e77", linewidth=1) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  # scale_y_continuous(breaks = sort(unique(error_two_curvy$b_non_empty))[-3]) +
  scale_x_continuous(breaks = sort(unique(round(error_two_curvy_umap$a1, 2)))[c(2, 4, 8, 11, 13, 14, 15, 16)]) +
  labs(x = expression(paste("binwidth (", a[1], ")")), y = expression(paste("proportion of non-empty bins ", bgroup("(", frac(m, b), ")")))) +
  ggtitle("(b)") +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 15),
axis.text = element_text(size = 12))
```

@@

```{r}
#| echo: false
#| fig-cap: "Various plots to help assess best hexagon bin parameters (a, b, d) and thresholds to remove low-density bins (c). Both (a) and (c) show MSE, against binwidth ($a_1$) and threshold. A good benchmark value for these parameters is when the MSE drops and then flattens out. Three binwidth choices were made: $0.03$, $0.05$, and $0.09$ to investigate. As the binwidth increases, the proportion of non-empty bins also increases (b). The relative proportion density decreases and levels off (d). Binwidth $0.05$ is chosen as the initial best binwidth for further analysis. There is no need to remove the low-density hexagons because as shown in (b), there is no considerable drop in MSE."
#| label: fig-param-two-curvy
#| out-width: 80%
#| fig-width: 12
#| fig-height: 10
#| fig-pos: H

@@

## hexagon binning to have regular hexagons
hb_obj_tsne_two_curvy2 <- hex_binning(
  data = tsne_two_curvy_scaled, 
  bin1 = 24, 
  r2 = r2_tsne)

a1_2_tsne <- calc_bins_y(
  bin1 = 24, 
  r2 = r2_tsne
)$a1
